\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[a4paper]{geometry}

\usepackage{amsthm}
\newtheorem{theo}{Theorem}
\newtheorem{defi}[theo]{Definition}
\newtheorem{prop}[theo]{Proposition}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{decorations.pathmorphing}

\setlength\parindent{0pt}

\newcommand{\La}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\ph}{\varphi}
\newcommand{\itemz}{\item[$\triangleright$]}
\newcommand{\F}{\mathcal{F}}
\newcommand{\gr}{\textbf}
\newcommand{\il}{\textit}
\newcommand{\N}{\mathbb{N}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\equi}{\Leftrightarrow}
\newcommand{\R}{\mathcal{R}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\iff}{\Leftrightarrow}
\newcommand{\T}{\mathbb{T}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\lb}{\llbracket}
\newcommand{\rb}{\rrbracket}
\newcommand{\info}[1]{\text{{\fontfamily{lmss}\selectfont #1}}}
\newcommand{\Mod}{\info{Mod}}
\newcommand{\Sen}{\info{Sen}}
\newcommand{\1}{\mathbbm{1}}
\renewcommand{\contentsname}{Table des matières}
\renewcommand{\proofname}{Preuve}
\renewcommand{\P}{\mathcal{P}}


\usepackage[bitstream-charter]{mathdesign}
%\usepackage[charter]{mathdesign}
%\usepackage{fontspec}
%\setmainfont{URW Palladio L}
\title{Category theory and data structures}


\begin{document}

\section{Category theory}
We all know that everything is a map, but category theory tells us more: everything is a commutative diagram.
\begin{defi}[Category]
A category $\gr{C}$ is composed of a set of objects and sets of morphisms. For each objects $A,B$, there is a set of morphisms from $A$ to $B$. Morphisms are denoted by $f : A \to B$. There is a composition law $\circ$ and for each object $A$ there is an identity morphism $id_A : A \to A$ such that
\begin{itemize}
\setlength\itemsep{-0.3em}
\itemz $\circ$ is associative: $h \circ (g \circ f) = (h \circ g) \circ f$.
\itemz $id_A$ is neutral: $f \circ id_A = id_B \circ f = f$.
\end{itemize}
\end{defi}

\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     A & B & C & D & & A & A \\
       &   &   &   & & B & B \\};
  \path[-stealth]
    (m-1-1) edge node [above] {$f$} (m-1-2)
    (m-1-2) edge [dashed] node [above] {$g$}(m-1-3)
    (m-1-3) edge node [above] {$h$} (m-1-4)
    (m-1-1) edge [bend left] node [above] {$g \circ f$} (m-1-3)
    (m-1-2) edge [bend right] node [below] {$h \circ g$} (m-1-4)
    (m-1-6) edge node [above] {$id_A$} (m-1-7)
    (m-1-7) edge node [right] {$f$} (m-2-7)
    (m-1-6) edge node [left] {$f$} (m-2-6)
    (m-2-6) edge node [below] {$id_B$} (m-2-7)
    (m-1-6) edge node [above right] {$f$} (m-2-7) ;
\end{tikzpicture}
\end{center}

\gr{Examples.}
\begin{itemize}
\setlength\itemsep{-0.3em}
\itemz $\gr{Set}$, objects are sets, morphisms are functions.
\itemz $\gr{Grp}$, objects are groups, morphisms are group homomorphisms.
\itemz $\gr{Vect}_K$ ($K$ is a field), objects are $K$-vector spaces, morphisms are $K$-linear applications.
\end{itemize}

\begin{defi}[Functor] Let $\gr{C}$ and $\gr{D}$ be categories. A functor $F : \gr{C} \to \gr{D}$ is a map that associates to each object $A \in \gr{C}$ an object $F(A) \in \gr{D}$, and to each morphism $f : A \to B$ in $\gr{C}$ a morphism $F(f) : F(A) \to F(B)$ in $\gr{D}$. A functor satisfies $F(g \circ f) = F(g) \circ F(f)$.
\end{defi}
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     FX &  \\
     FY & FZ \\};
  \path[-stealth]
    (m-1-1) edge node [left] {$F(f)$} (m-2-1)
    (m-2-1) edge node [below] {$F(g)$}(m-2-2)
    (m-1-1) edge node [above right] {$F(g\circ f)$} (m-2-2);
\end{tikzpicture}
\end{center}

\gr{Example.} \begin{itemize}
\itemz In $\gr{Set}$, the powerset functor $\P$ associates to a set $A$ the set $\P(A)$ of all sets included in $A$, and to each function $f : A \to B$ the direct image function $\P(f) : \P(A) \to \P(B)$ defined by $\P(f)(U) = \{ f(u) \mid u \in U \}$. (Homework: Check that it is a functor!)
\end{itemize}
\begin{defi}[Initial object]
Let $\gr{C}$ be a category and $A$ an object. $I$ is an initial object if for any object $A$ there exists exactly one morphism $\lb A \rb : I \to A$.
\end{defi}
\newpage
\section{$F$-algebras}
Now we know what is a category and what is a functor. These are the only tools we need in order to formalize properly mathematical induction, through what is called $F$-algebras.
\begin{defi}
Let $\gr{C}$ be a category and and $F : \gr{C} \to \gr{C}$ be a functor. (Note that this is an endofunctor: the domain category is the same as the codomain category.) An $F$-algebra is a morphism $\alpha : F(A) \to A$, where $A$ is an object of $\gr{C}$ called the carrier of $\alpha$.
\end{defi}
\begin{defi} Let $\alpha : F(A) \to A$ and $\beta : F(B) \to B$ be two $F$-algebras. A homomorphism of $F$-algebras from $\alpha$ to $\beta$ is a morphism $h : A \to B$ such that $\beta \circ F(h) = h \circ \alpha$.
\end{defi}
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     F(A) & A \\
     F(B) & B \\};
  \path[-stealth]
    (m-1-1) edge node [left] {$F(h)$} (m-2-1)
    (m-2-1) edge node [below] {$\beta$}(m-2-2)
    (m-1-1) edge node [above] {$\alpha$} (m-1-2)
    (m-1-2) edge node [right] {$h$} (m-2-2);
\end{tikzpicture}
\end{center}
\begin{prop}
The $F$-algebras and their homomorphisms form a category denoted by $\gr{Alg}(F)$.
\end{prop}
\begin{proof}[Proof] Homework. Let us just prove that composition defines a $F$-algebra homomorphism. Proof by diagram:
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     F(A) & A \\
     F(B) & B \\
     F(C) & C \\};
  \path[-stealth]
    (m-1-1) edge node [right] {$F(h)$} (m-2-1)
    (m-2-1) edge node [below] {$\beta$}(m-2-2)
    (m-1-1) edge node [above] {$\alpha$} (m-1-2)
    (m-1-2) edge node [left] {$h$} (m-2-2)
    (m-2-1) edge node [right] {$F(k)$} (m-3-1)
    (m-3-1) edge node [below] {$\gamma$} (m-3-2)
    (m-2-2) edge node [left] {$k$} (m-3-2)
    (m-1-1) edge [bend right] node [left] {$F(k \circ h)$} (m-3-1)
    (m-1-2) edge [bend left] node [right] {$k\circ h$} (m-3-2);
\end{tikzpicture}
\end{center}
\end{proof}
If there exists an initial object in $\gr{Alg}(F)$, then it is of great importance. Unfold the definitions to see what it means to be an initial object in $\gr{Alg}$.\\\\
$\iota : F(I) \to I$ is initial if for any other $\alpha : F(A) \to A$ there exists a unique morphism $\lb \alpha \rb : I \to A$ such that the following diagram commutes.
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     F(I) & I \\
     F(A) & A \\};
  \path[-stealth]
    (m-1-1) edge node [left] {$F(\lb \alpha \rb)$} (m-2-1)
    (m-2-1) edge node [below] {$\alpha$}(m-2-2)
    (m-1-1) edge node [above] {$\iota$} (m-1-2)
    (m-1-2) edge node [right] {$\lb \alpha \rb$} (m-2-2);
\end{tikzpicture}
\end{center} 
\newpage
\section{Initial algebras in computer science}
One can prove that if there exists an initial algebra, then it is unique (up-to unique isomorphism, but let us forget this). Initial algebras are extremly useful in computer science. Everybody implicitly uses them when writing recursive functions. Let us give three examples. All of them are living in the category $\gr{Set}$.
\begin{itemize}
\itemz Natural numbers.
\itemz Lists.
\itemz Binary trees.
\end{itemize}
\subsection*{Natural numbers}
According to the Peano axioms, there are two ways of obtaining a natural number. Either you take $0$, either you take the successor of some natural number $n$, written $n+1$. This is formalized by the following functor.
\begin{defi}
The functor $F : \gr{Set} \to \gr{Set}$ is defined by $F(A) = 1 + A$ the disjoint union of $A$ and a set $1$ with only one element denoted by $1 = \{*\}$. If $f : A \to B$ is a function then $F(f) = id_1 + f$ is the function defined by $F(f)(*) = *$ and $F(f)(a) = f(a)$ for any $a \in A$.
\end{defi}
In order to define a function on $1 + A$, we have to specify what is its value on $*$ and on every element of $A$. For example:
\begin{defi} $\iota : 1 + \N \to \N$ is defined by $\iota(*) = 0$ and $\iota(n) = n+1$. \end{defi}
\begin{theo} $\iota$ is the initial $F$-algebra. \end{theo} 
This means that for every morphism $\alpha : 1 + A \to A$, there is a unique morphism $\lb \alpha \rb : \N \to A$ such that the following diagram commutes.
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     1 + \N & \N \\
     1 + A & A \\};
  \path[-stealth]
    (m-1-1) edge node [left] {$id_1 + \lb \alpha \rb$} (m-2-1)
    (m-2-1) edge node [below] {$\alpha$}(m-2-2)
    (m-1-1) edge node [above] {$\iota$} (m-1-2)
    (m-1-2) edge node [right] {$\lb \alpha \rb$} (m-2-2);
\end{tikzpicture}
\end{center} 
Looking at the diagram, one sees that
\begin{itemize}
\setlength\itemsep{-0.3em}
\itemz $\lb \alpha \rb (0) = \alpha(*)$
\itemz $\lb \alpha \rb (n+1) = \alpha(\lb \alpha \rb (n))$
\end{itemize}
Hence $\lb \alpha \rb(n) = \alpha^n(*)$. The morphism $\lb \alpha \rb$ is formalizing the type of induction that natural numbers stand for, i.e., one-directional, repeated application of the same function.
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     0 & 1 & 2 & 3 & 4 & 5 & \text{etc.} \\};
  \path[-stealth]
    (m-1-1) edge node [above] {$+1$} (m-1-2)
    (m-1-2) edge node [above] {$+1$} (m-1-3)
    (m-1-3) edge node [above] {$+1$} (m-1-4)
    (m-1-4) edge node [above] {$+1$} (m-1-5)
    (m-1-5) edge node [above] {$+1$} (m-1-6)
    (m-1-6) edge node [above] {$+1$} (m-1-7);
\end{tikzpicture}
\end{center} 
\newpage
\subsection*{Lists}
Remember what a list is in computer science. First, it is a list \il{of something}. We choose to look at lists of \il{characters}, i.e., each element a list is a character living in an alphabet $\Sigma$. How do we define a list of characters in computer science ? Either it is the empty list, denoted by $[]$, or it is the concatenation of one character $c$ and a list $l$, denoted by $c::l$. In your opinion, what will be the functor that will be useful to model this situation?
\begin{defi}
The functor $F : \gr{Set} \to \gr{Set}$ is defined by $F(A) = 1 + \Sigma \times A$ the disjoint union of the set of pairs $\Sigma \times A$ and $1 = \{*\}$. If $f : A \to B$ is a function then $F(f) = id_1 + id_\Sigma \times f$ is the function defined by $F(f)(*) = *$ and $F(f)((c,a)) = (c,f(a))$ for any $(c,a) \in \Sigma \times A$.
\end{defi}
In order to define a function on $1 + \Sigma \times A$, we have to specify what is its value on $*$ and on every element of $\Sigma \times A$. Let us take $A = \Sigma^*$ the set of lists of characters in $\Sigma$ (for example, if $\Sigma = \{a,b\}$, $\Sigma^* = \{[], [a], [b], [a,b], [a,a], [b,b], [b,a], [a,a,b], [a,b,a], [a,a,b,a,b,b,b]...\}$). Then we define the following $F$-algebra.
\begin{defi} $\iota : 1 + \Sigma \times \Sigma^* \to \Sigma^*$ is defined by $\iota(*) = []$ (the empty list) and $\iota((c,l)) = c::l$.\end{defi}
\begin{theo} $\iota$ is the initial $F$-algebra. \end{theo} 
This means that for every morphism $\alpha : 1 + \Sigma \times A \to A$, there is a unique morphism $\lb \alpha \rb : \Sigma^* \to A$ such that the following diagram commutes.
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     1 + \Sigma \times \Sigma^* & \Sigma^* \\
     1 + \Sigma \times A & A \\};
  \path[-stealth]
    (m-1-1) edge node [left] {$id_1 + id_\Sigma \times \lb \alpha \rb$} (m-2-1)
    (m-2-1) edge node [below] {$\alpha$}(m-2-2)
    (m-1-1) edge node [above] {$\iota$} (m-1-2)
    (m-1-2) edge node [right] {$\lb \alpha \rb$} (m-2-2);
\end{tikzpicture}
\end{center} 
Looking at the diagram, one sees that
\begin{itemize}
\setlength\itemsep{-0.3em}
\itemz $\lb \alpha \rb ([]) = \alpha(*)$
\itemz $\lb \alpha \rb (c::l) = \alpha((c,\lb \alpha \rb(l)))$
\end{itemize}
There is no simple expression for $\lb \alpha \rb$. The morphism $\lb \alpha \rb$ is formalizing the type of induction that lists stand for. The fact that the initial $F$-algebra is precisely the algebra of lists means that lists are a very canonical structure. 
\newpage
\subsection*{Binary trees}
A binary tree $T$ is either a root $\bullet$ or composed of a root $\bullet$ and two  binary trees $L,R$ ($L$ is the left sub-tree and $R$ is the right sub-tree). It is denoted by $T(L,R)$ and represented by
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
     & \bullet & \\
     L & & R \\};
  \path[-]
    (m-1-2) edge node [above left] {} (m-2-1)
    (m-1-2) edge node [above right] {} (m-2-3);
\end{tikzpicture}
\end{center} 
\begin{defi}
The functor $F : \gr{Set} \to \gr{Set}$ is defined by $F(A) = 1 + A \times A$. If $f : A \to B$ is a function then $F(f) = id_1 + f \times f$ is the function defined by $F(f)(*) = *$ and $F(f)((a,b)) = (f(a),f(b))$ for any $(a,b) \in A \times A$.
\end{defi}
Let take as $A$ the set $\T$ of all binary trees. Then we can define
\begin{defi} $\iota : 1 + \T \times \T \to \T$ is defined by $\iota(*) = \bullet$ and $\iota((L,R))$ is the tree $T(L,R)$.
\end{defi}


\end{document}